TrueSkill Through Time: the \proglang{Julia}, \proglang{Python} and \proglang{R} packages

### 2021-03-29 article.tex
TTT: 0 warnings

\usepackage{wrapfig}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{amsmath} %para escribir funci\'on partida , matrices
\usepackage{amsthm} %para numerar definciones y teoremas
\usepackage{amsfonts} % \mathbb{N} -> conjunto de los n\'umeros naturales
\usepackage{bm} % \bm{\alpha} bold greek symbol
\usepackage[makeroom]{cancel} % \cancel{} \bcancel{} etc
\usepackage{mdframed}
\usepackage{algorithm}
\usepackage{comment}
%\usepackage{quoting}
\usepackage{mathtools}
\usepackage{tikz}
\usepackage{csvsimple}
\usepackage{listings}
\renewcommand{\lstlistingname}{Code}% Listing -> Algorithm
\usepackage{hyperref}
\usepackage{todonotes}
\usepackage[binary-units]{siunitx}

@@

\en{All currently used skill estimators share some variant of the probabilistic model proposed by Elo~\citep{Glickman1999,Herbrich2007,VanDerLinden2016,Fox2010}.}
\es{Todos los estimadores de habilidad ampliamente utilizados actualmente comparten alguna variante del modelo probabil\'istico propuesto por Elo~\cite{Glickman1999, Herbrich2007, VanDerLinden2016, Fox2010}.}
%
\begin{figure}[h!]
\centering \small
    \tikz{         
    \node[det, fill=black!10] (r) {$r$} ; 
    \node[const, left=of r, xshift=-1.35cm] (r_name) {\small \en{Result}\es{Resultado}:}; 
    \node[const, right=of r] (dr) {\normalsize $ r = (d > 0)$};

@@

% \begin{figure}[h!]

@@

\en{TrueSkill Through Time outperforms TrueSkill and others filtering models.}
\es{TrueSkill Through Time supera a TrueSkill y otros modelos de filtrado.}
%
\en{In Figure~\ref{fig:smooth_example} we show the estimation's behavior in a data set with two agents and two events.}
\es{En la figura~\ref{fig:smooth_example} se muestra el comportamiento de las estimaciones en conjunto de datos de dos agentes y dos eventos.}
\begin{figure}[h!]
  \centering
  \scalebox{.9}{
    \tikz{ %
      \node[latent] (s10) {$s_{a_0}$} ;
      %
      \node[latent,  below=of s10,yshift=-1cm] (s11) {$s_{a_1}$} ;

@@

    \en{\includegraphics[page={1},width=.75\linewidth]{figures/optimization.pdf}}
    \es{\includegraphics[page={1},width=.75\linewidth]{figures/optimization.pdf}}

@@

\en{In this last example we use the entire history of matches of the Association of Tennis Professionals (ATP).}
\es{En este último ejemplo utilizamos toda la historia de partidas de la Asociación de Tenistas Profesionales (ATP).}
%
\en{The database has \num{447000} games starting at year 1915 until 2020 with more than \num{19000} participating players and is publicly available\footnote{A zip file containing a csv file per year can be downloaded from \url{https://github.com/JeffSackmann/tennis_atp}}.}
\es{La base de datos cuenta con \num{447000} mil partidas que van desde el año 1915 hasta el 2020, en las que participan \num{19000} jugadores.}
%
\en{The information is formatted and stored in a single \texttt{csv} file\footnote{Available at \url{https://github.com/glandfried/tennis_atp/releases/download/atp/history.csv.zip}} including both single and double matches: if the column \texttt{double} has the letter \texttt{t}, the game is a double match.}
\es{La información contenida en los archivos \texttt{csv} contiene tanto partidas simples como dobles, indicado por la columna \texttt{double}.}
%
\en{Each game has a unique identifier (i.e. \texttt{match\_id}) and tournament's round number (i.e. \texttt{round\_number}) where 0 represents the final game, 1 the semi-final, and so on.}
\es{Tenemos un identificador por la partida (\texttt{match\_id}) y el número de ronda (\texttt{round\_number}) donde 0 representa la partida final, 1 la semi-final, etc.}
%
\en{The file also contains four columns for the players' identifiers in the game (the last two players are null in the case of a single): for example \texttt{w2\_id} is the identifier of the second player of the winning team and \texttt{l1\_name} is the name of the first player of the losing team.}
\es{Contiene cuatro columnas correspondiente a los identificadores de los jugadores y cuatro columnas correspondiente a sus nombres, por ejemplo \texttt{w2\_id} es el identificador del segundo jugador del equipo ganador (los segundos jugadores tienen valores nulos en partidas simples) y \texttt{l1\_name} es el nombre del primer jugador del equipo perdedor.}
%
\en{Finally, we have the name of the tournament (\texttt{tour\_name}), its identifier (\texttt{tour\_id}), the start date of the tournament (\texttt{time\_start}) and the type of surface (\texttt{ground}).}
\es{Finalmente tenemos el nombre del torneo (\texttt{tour\_name}), su identificado (\texttt{tour\_id}), la fecha de inicio del torneo (\texttt{time\_start}) y el tipo de piso (\texttt{ground}).}
%
\en{Here we only show the \proglang{Julia} version because it is more efficient than \proglang{Python} and \proglang{R} versions, which can be also found in the appendix \ref{AGREGA-CODE}.}
\es{Aquí mostramos solo la versión de \proglang{Julia} debido a que es la más eficiente, las versiones de \proglang{Python} y \proglang{R} se pueden entonrar en el anexo \ref{AGREGA-CODIGO}.}
%
\begin{lstlisting}[backgroundcolor=\color{julia},label=lst:atp, caption={\en{The history of the Association of Tennis Professionals}\es{Historia sintética}}, belowskip=0.1cm, aboveskip=0.1cm]
using CSV; using Dates
data = CSV.read("atp.csv")

@@

h = History(composition=composition, times = times, sigma = 2.0, gamma = 0.02)
convergence(h,epsilon=0.01, iterations=10)
\end{lstlisting}
%
\en{In this code we open the file \texttt{atp.csv}, create the variables \texttt{times} and \texttt{composition}, and instantiate the class \texttt{History}.}
\es{En este código abrimos el archivo \texttt{csv}, creamos las variables \texttt{times} y \texttt{composition} e instanciamos la clase \texttt{History}.}
%
\en{Temporal batches are defined as days elapsed since a reference date to the tournament start.}
\es{Los lotes temporales se definen como días transcurridos desde una fecha de referencia hasta el inicio del torneo.}
%
\en{When generating the list \texttt{composition} we discriminate whether the games are double or single based on the column \texttt{double}.}
\es{Al generar la lista \texttt{composition} discriminamos si los juegos son dobles o individuales en base a la columna \texttt{double}.}
%
\en{The results are implied by the order in which we set the composition, always placing the winning team first.}
\es{Los resultados quedan impl\'icitos por el orden en que armamos la composición, poniendo siempre primero al equipo ganador.}
%
\en{When initializing the class \texttt{History} we also modify the default values of \texttt{sigma} and \texttt{gamma} based on an optimization procedure previously performed (more details next).}
\es{Al inicializar la clase \texttt{History} modificamos los valores por defecto de \texttt{sigma} y \texttt{gamma}, basados en un procedimiento de optimización realizado previamente (más detalles a continuación).}
%
\en{Finally, we use the \texttt{convergence()} method to obtain TrueSkill Through Time estimates explicitly selecting the convergence criterion: when the change between iterations is less than $0.01$ or when $10$ iterations are performed.}
\es{Finalmente, convergemos la historia para obtener las estimaciones de TrueSkill Through Time, indicando explícitamente el criterio de corte: cuando el cambio entre iteraciones sea menor a $0.01$, o cuando se alcanzan las $10$ iteraciones.}
%
\en{Figure~\ref{fig:atp} presents the learning curves of eight famous players of the ATP's history.}
\es{En la figura~\ref{fig:atp} graficamos las curvas de aprendizaje de ocho jugadores famosos de la historia de la ATP.}
%
\begin{figure}[h]
    \centering
    \includegraphics[page={1},width=.7\linewidth]{figures/atp}
    \caption{ATP}
    \label{fig:atp}
\end{figure}

@@

\en{The \emph{sum-product algorithm}~\citep{Kschischang2001} is a general procedure that takes advantage of the structure of the joint probability distribution imposed by the causal model to efficiently apply the rules of probability, the \ref{eq:sum_rule} and the \ref{eq:product_rule}.}
\es{El \emph{sum-product algorithm}~\citep{Kschischang2001} es un procedimiento general que aprovecha la estructura de la disstribuci\'on de probabilidad conjunta que impone el modelo causal para aplicar eficientemente las reglas de la probabilidad, la~\ref{eq:sum_rule} y la~\ref{eq:product_rule}.}
%
\en{Any model can be factored into the product of conditional probabilities.}
\es{Cualquier modelo puede factorizarse en el producto de probabilidades condicionales.}
%
\en{By making use of the independencies between variables (figure~\ref{fig:generative_model}) our model can be factored as,}
\es{Haciendo uso de las independencias entre las variables (figura~\ref{fig:generative_model}) nuestro modelo puede factorizarse como,}
%
\begin{equation} \label{eq:factorization}
 p(\bm{s},\bm{p},d,r) = p(s_1)p(s_2)p(p_1|s_1)p(p_2|s_2)p(d|\bm{p})P(r|d)
\end{equation}
%
\en{In figure~\ref{fig:factor_graph} we show this factorization graphically.}
\es{En la figura~\ref{fig:factor_graph} mostramos gr\'aficamente la factorizaci\'on.}
%
\en{These types of representations, known as \emph{factor graph}, are graphs with two types of nodes: variable nodes (white circles), and function nodes (black squares).}
\es{A este tipo de representaciones, conocidas como \emph{factor graph}, son gr\'afos con dos tipos de nodos: los nodos variables (círculos blancos), y los nodos funciones (cuadrados negros).}
%
\en{The edge between node variables and node functions represent the mathematical relationship ``the variable $v$ is an argument of the function $f$''.}
\es{Los ejes entre los nodos variables y los nodos funciones representan la relaci\'on matem\'atica ``la variable $v$ es argumento de la funci\'on $f$''.}
%
\begin{figure}[h!]
\centering \small
    \tikz{         
%         \node[const, above=of fr] (nfr) {$f_r$}; %
% 	\node[const, above=of nfr] (dfr) {\large $\mathbb{I}(d >0)$}; %

@@

\begin{equation}\label{eq:m_v_f} \tag{\text{product step}}

@@

\begin{equation}\label{eq:m_f_v}  \tag{\text{sum step}}

@@

\begin{equation}\label{eq:marginal}
g_i(x_i) = \prod_{h \in n(x_i)} m_{h \rightarrow x_i}

@@

\begin{equation}\label{eq:gaussian_product} \tag{\text{\en{Gaussian product}\es{Producto de gaussianas}}}

@@

\begin{equation}\label{eq:gaussian_division} \tag{\text{\en{Gaussian division}\es{División de gaussianas}}}

@@

\begin{equation}\label{eq:integral_con_dirac} \tag{\text{\en{Dirac delta function}\es{Función delta de dirac}}}

@@

\begin{equation}\label{eq:simetria} \tag{\text{\en{Gaussian symmentry}\es{Simetría de gaussianas}}}

@@

\begin{equation}\label{eq:estandarizar} \tag{\text{\en{Gaussian standarization}\es{Estandarización de gaussianas}}}

@@

\begin{equation}\label{eq:phi_norm} \tag{\text{\en{Derivative of the cumulative Gaussian}\es{Derivada de la gaussiana acumulada}}}

@@

\begin{equation}\label{eq:phi_simetria} \tag{\text{\en{Symmetry of the cumulative Gaussian}\es{Simetría de la gaussiana acumulada}}}

@@

\en{The \texttt{Gaussian} class does most of the computation of the packages.}
\es{La clase \texttt{Gaussian} realiza la mayor parte del c\'omputo en todos los paquetes.}
%
\en{It is represented by two parameters, the mean (\texttt{mu}) and the standard deviation (\texttt{sigma}).}
\es{Se representa mediante dos par\'ametros, la media y el desv\'io estandar.}
% %
\begin{lstlisting}[backgroundcolor=\color{all},label=lst:N1_N2, caption={\en{Initialization of Gaussians distributions}\es{Inicialización de distirbuciones gaussianas}}, belowskip=-1.0 \baselineskip, aboveskip=-0 \baselineskip]
\end{lstlisting}
\begin{lstlisting}[backgroundcolor=\color{all}, belowskip=0.0 \baselineskip]
N1 = Gaussian(mu = 1.0, sigma = 1.0); N2 = Gaussian(1.0, 2.0)  
\end{lstlisting}
%
\en{The class overwrites the operators addition (\texttt{+}), subtraction (\texttt{-}), product (\texttt{*}) and division (\texttt{/}) with the main properties required to compute the marginal distributions in the TrueSkill Through Time model.}
\es{La clase sobreescribe los operadores suma (\texttt{+}), resta (\texttt{-}), producto (\texttt{*}) y divisi\'on (\texttt{/}) con las principales propiedades requeridas para computar las distribuciones marginales en el modelo TrueSkill Through Time.}
%
\begin{equation} \tag{\texttt{N1 * N2}}
 \N(x|\mu_1,\sigma_1^2)\N(x|\mu_2,\sigma_2^2) \overset{\ref{multiplicacion_normales}}{\propto} \N(x|\mu_{*},\sigma_{*}^2)
\end{equation}
%
\begin{equation} \tag{\texttt{N1 / N2}}
 \N(x|\mu_1,\sigma_1^2)/\N(x|\mu_2,\sigma_2^2)  \overset{\ref{sec:division_normales}}{\propto} \N(x|\mu_{\div},\sigma_{\div}^2)
\end{equation} 
%
\vspace{-0.3cm}
%
\begin{equation} \tag{\texttt{N1 + N2}} \label{eq:suma_normales}
\begin{split}
\iint \delta(t=x + y) \N(x|\mu_1, \sigma_1^2)\N(y|\mu_2, \sigma_2^2) dxdy \overset{\text{\ref{suma_normales_induccion}}}{=} \N(t|\mu_1+\mu_2,\sigma_1^2 + \sigma_2^2)
\end{split}
\end{equation}
%
\vspace{-0.5cm}
%
\begin{equation} \tag{\texttt{N1 - N2}} \label{eq:resta_normales}
\begin{split}
\iint \delta(t = x - y) \N(x|\mu_1, \sigma_1^2)\N(y|\mu_2, \sigma_2^2) dxdy \overset{\text{\ref{suma_normales_induccion}}}{=} \N(t|\mu_1 - \mu_2,\sigma_1^2 + \sigma_2^2)
\end{split}
\end{equation}

@@

\en{TrueSkill allows to estimate the skill of individuals even when they play in teams.}
\es{TrueSkill permite estimar la habilidad de los individuos incluso cuando estos juegan en equipo.}
%
\en{The model assumes that the team with the highest performance wins, $r = (t_i > t_j)$, where $t_i$ and $t_j$ represent the team performances.}
\es{El modelo supone que gana el equipo con mayor rendimiento, $r = (t_i > t_j)$, donde $t_i$ y $t_j$ representan los desempeños de los equipos.}
%
\en{In figure~\ref{fig:modelo_trueskill_2vs2} we show the factor graph.}
\es{En la figura~\ref{fig:modelo_trueskill_2vs2} mostramos la factorizaci\'on gr\'afica.} 
%
\begin{figure}[h!]
  \centering
  \scalebox{.9}{
  \tikz{

@@

\en{Following the \ref{eq:m_f_v} of the sum-product algorithm and the factorization of the model displayed at Figure~\ref{fig:modelo_trueskill_2vs2}, the messages from the skill factors $f_{s_i}$ to their variable $s_i$ are just the priors.}
\es{Siguiendo la regla~\ref{eq:m_f_v} del \emph{sum-product algorithm} y la factorizaci\'on del modelo presentado en la figura~\ref{fig:modelo_trueskill_2vs2}, los mensajes de los factores de habilidad $f_{s_i}$ a su variable $s_i$ no son otra cosa m\'as que el priors.}
\begin{equation}\label{eq:m_fs_s} \tag{\texttt{prior}}

@@

\en{The next message, the one that the variable $s_i$ sends to the performance factor $f_{p_i}$ is also the prior, in this case due to the~\ref{eq:m_v_f} of the sum-product algorithm and the factorization of the model.}
\es{El siguiente mensaje, el que la variable $s_i$ le env\'ia al factor rendimiento $f_{p_i}$ tambi\'en es el prior, en este caso debido al~\ref{eq:m_v_f} del \emph{sum-product algorithm} y la factorizacion del modelo.}

@@

\begin{equation}\label{eq:m_fp_p} \tag{\texttt{performance()}}

@@

\begin{equation} \label{eq:m_ft_t} \tag{\texttt{ta = p1 + p2}}

@@

\begin{equation}\label{eq:m_fd_d} \tag{\texttt{d = ta - tb}}

@@

\begin{equation}\label{eq:m_fr_r} \tag{\texttt{evidence}}

@@

\begin{figure}[h!]

@@

@@

% Appendices can be included after the bibliography (with a page break). Each
% section within the appendix should have a proper section title (rather than
% just \emph{Appendix}).

@@

\newpage
\end{document}

>>>>

\usepackage{wrapfig}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{amsmath} %para escribir funci\'on partida , matrices
\usepackage{amsthm} %para numerar definciones y teoremas
\usepackage{amsfonts} % \mathbb{N} -> conjunto de los n\'umeros naturales
\usepackage{bm} % \bm{\alpha} bold greek symbol
\usepackage[makeroom]{cancel} % \cancel{} \bcancel{} etc
\usepackage{mdframed}
\usepackage{algorithm}
\usepackage{comment}
%\usepackage{quoting}
\usepackage{mathtools}
\usepackage{tikz}
\usepackage{csvsimple}
\usepackage{listings}
\renewcommand{\lstlistingname}{Code}% Listing -> Algorithm
\usepackage{todonotes}
\usepackage[binary-units]{siunitx}

 % by: https://tex.stackexchange.com/questions/87940/weird-warning-using-pdfx
\RequirePackage{pdf14}

% by: https://tex.stackexchange.com/questions/74636/mla-package-and-thumbpdf
\makeatletter
\@namedef{ver@thumbpdf.sty}{}
\makeatother

@@

\en{All currently used skill estimators share some variant of the probabilistic model proposed by Elo~\citep{Glickman1999,Herbrich2007,VanDerLinden2016,Fox2010}.}
\es{Todos los estimadores de habilidad ampliamente utilizados actualmente comparten alguna variante del modelo probabil\'istico propuesto por Elo~\cite{Glickman1999, Herbrich2007, VanDerLinden2016, Fox2010}.}
%
\begin{figure}[ht!]
\centering \small
    \tikz{         
    \node[det, fill=black!10] (r) {$r$} ; 
    \node[const, left=of r, xshift=-1.35cm] (r_name) {\small \en{Result}\es{Resultado}:}; 
    \node[const, right=of r] (dr) {\normalsize $ r = (d > 0)$};

@@

% \begin{figure}[ht!]

@@

\en{TrueSkill Through Time outperforms TrueSkill and others filtering models.}
\es{TrueSkill Through Time supera a TrueSkill y otros modelos de filtrado.}
%
\en{In Figure~\ref{fig:smooth_example} we show the estimation's behavior in a data set with two agents and two events.}
\es{En la figura~\ref{fig:smooth_example} se muestra el comportamiento de las estimaciones en conjunto de datos de dos agentes y dos eventos.}
\begin{figure}[ht!]
  \centering
  \scalebox{.9}{
    \tikz{ %
      \node[latent] (s10) {$s_{a_0}$} ;
      %
      \node[latent,  below=of s10,yshift=-1cm] (s11) {$s_{a_1}$} ;

@@

    \en{\includegraphics[page={1},width=.7\linewidth]{figures/optimization.pdf}}
    \es{\includegraphics[page={1},width=.7\linewidth]{figures/optimization.pdf}}

@@

\en{In this last example we use the entire history of matches of the Association of Tennis Professionals (ATP).}
\es{En este último ejemplo utilizamos toda la historia de partidas de la Asociación de Tenistas Profesionales (ATP).}
%
\en{The database has \num{447000} games starting at year 1915 until 2020 with more than \num{19000} participating players and is publicly available\footnote{A zip file containing a csv file per year can be downloaded from \nolinkurl{https://github.com/JeffSackmann/tennis_atp}}.}
\es{La base de datos cuenta con \num{447000} mil partidas que van desde el año 1915 hasta el 2020, en las que participan \num{19000} jugadores.}
%
\en{The information is formatted and stored in a single \texttt{csv} file\footnote{Available at \nolinkurl{https://github.com/glandfried/tennis_atp/releases/download/atp/history.csv.zip}} including both single and double matches: if the column \texttt{double} has the letter \texttt{t}, the game is a double match.}
\es{La información contenida en los archivos \texttt{csv} contiene tanto partidas simples como dobles, indicado por la columna \texttt{double}.}
%
\en{Each game has a unique identifier (i.e. \texttt{match\_id}) and tournament's round number (i.e. \texttt{round\_number}) where 0 represents the final game, 1 the semi-final, and so on.}
\es{Tenemos un identificador por la partida (\texttt{match\_id}) y el número de ronda (\texttt{round\_number}) donde 0 representa la partida final, 1 la semi-final, etc.}
%
\en{The file also contains four columns for the players' identifiers in the game (the last two players are null in the case of a single): for example \texttt{w2\_id} is the identifier of the second player of the winning team and \texttt{l1\_name} is the name of the first player of the losing team.}
\es{Contiene cuatro columnas correspondiente a los identificadores de los jugadores y cuatro columnas correspondiente a sus nombres, por ejemplo \texttt{w2\_id} es el identificador del segundo jugador del equipo ganador (los segundos jugadores tienen valores nulos en partidas simples) y \texttt{l1\_name} es el nombre del primer jugador del equipo perdedor.}
%
\en{Finally, we have the name of the tournament (\texttt{tour\_name}), its identifier (\texttt{tour\_id}), the start date of the tournament (\texttt{time\_start}) and the type of surface (\texttt{ground}).}
\es{Finalmente tenemos el nombre del torneo (\texttt{tour\_name}), su identificado (\texttt{tour\_id}), la fecha de inicio del torneo (\texttt{time\_start}) y el tipo de piso (\texttt{ground}).}
%
\en{Here we only show the \proglang{Julia} version because it is more efficient than \proglang{Python} and \proglang{R} versions, which can be also found in the appendix \ref{sec:appendix_atp_code}.}
\es{Aquí mostramos solo la versión de \proglang{Julia} debido a que es la más eficiente, las versiones de \proglang{Python} y \proglang{R} se pueden entonrar en el anexo \ref{sec:appendix_atp_code}.}
%
\begin{lstlisting}[backgroundcolor=\color{julia},label=lst:atp, caption={\en{The history of the Association of Tennis Professionals}\es{Historia sintética}}, belowskip=0.1cm, aboveskip=0.1cm]
using CSV; using Dates
data = CSV.read("atp.csv")

@@

h = History(composition=composition, times = times, sigma = 2.0, gamma = 0.02)
convergence(h,epsilon=0.01, iterations=10)
\end{lstlisting}
%
\en{In this code we open the file \texttt{atp.csv}, create the variables \texttt{times} and \texttt{composition}, and instantiate the class \texttt{History}.}
\es{En este código abrimos el archivo \texttt{csv}, creamos las variables \texttt{times} y \texttt{composition} e instanciamos la clase \texttt{History}.}
%
\en{Temporal batches are defined as days elapsed since a reference date to the tournament start.}
\es{Los lotes temporales se definen como días transcurridos desde una fecha de referencia hasta el inicio del torneo.}
%
\en{When generating the list \texttt{composition} we discriminate whether the games are double or single based on the column \texttt{double}.}
\es{Al generar la lista \texttt{composition} discriminamos si los juegos son dobles o individuales en base a la columna \texttt{double}.}
%
\en{The results are implied by the order in which we set the composition, always placing the winning team first.}
\es{Los resultados quedan impl\'icitos por el orden en que armamos la composición, poniendo siempre primero al equipo ganador.}
%
\en{When initializing the class \texttt{History} we also modify the default values of \texttt{sigma} and \texttt{gamma} based on an optimization procedure previously performed (more details next).}
\es{Al inicializar la clase \texttt{History} modificamos los valores por defecto de \texttt{sigma} y \texttt{gamma}, basados en un procedimiento de optimización realizado previamente (más detalles a continuación).}
%
\en{Finally, we use the \texttt{convergence()} method to obtain TrueSkill Through Time estimates explicitly selecting the convergence criterion: when the change between iterations is less than $0.01$ or when $10$ iterations are performed.}
\es{Finalmente, convergemos la historia para obtener las estimaciones de TrueSkill Through Time, indicando explícitamente el criterio de corte: cuando el cambio entre iteraciones sea menor a $0.01$, o cuando se alcanzan las $10$ iteraciones.}
%
\en{Figure~\ref{fig:atp} presents the learning curves of eight famous players of the ATP's history.}
\es{En la figura~\ref{fig:atp} graficamos las curvas de aprendizaje de ocho jugadores famosos de la historia de la ATP.}
%
\begin{figure}[ht!]
    \centering
    \includegraphics[page={1},width=.7\linewidth]{figures/atp}
    \caption{ATP}
    \label{fig:atp}
\end{figure}

@@

\en{The \emph{sum-product algorithm}~\citep{Kschischang2001} is a general procedure that takes advantage of the structure of the joint probability distribution imposed by the causal model to efficiently apply the rules of probability, the \ref{eq:sum_rule} and the \ref{eq:product_rule}.}
\es{El \emph{sum-product algorithm}~\citep{Kschischang2001} es un procedimiento general que aprovecha la estructura de la disstribuci\'on de probabilidad conjunta que impone el modelo causal para aplicar eficientemente las reglas de la probabilidad, la~\ref{eq:sum_rule} y la~\ref{eq:product_rule}.}
%
\en{Any model can be factored into the product of conditional probabilities.}
\es{Cualquier modelo puede factorizarse en el producto de probabilidades condicionales.}
%
\en{By making use of the independencies between variables (figure~\ref{fig:generative_model}) our model can be factored as,}
\es{Haciendo uso de las independencias entre las variables (figura~\ref{fig:generative_model}) nuestro modelo puede factorizarse como,}
%
\begin{equation} \label{eq:factorization}
 p(\bm{s},\bm{p},d,r) = p(s_1)p(s_2)p(p_1|s_1)p(p_2|s_2)p(d|\bm{p})P(r|d)
\end{equation}
%
\en{In figure~\ref{fig:factor_graph} we show this factorization graphically.}
\es{En la figura~\ref{fig:factor_graph} mostramos gr\'aficamente la factorizaci\'on.}
%
\en{These types of representations, known as \emph{factor graph}, are graphs with two types of nodes: variable nodes (white circles), and function nodes (black squares).}
\es{A este tipo de representaciones, conocidas como \emph{factor graph}, son gr\'afos con dos tipos de nodos: los nodos variables (círculos blancos), y los nodos funciones (cuadrados negros).}
%
\en{The edge between node variables and node functions represent the mathematical relationship ``the variable $v$ is an argument of the function $f$''.}
\es{Los ejes entre los nodos variables y los nodos funciones representan la relaci\'on matem\'atica ``la variable $v$ es argumento de la funci\'on $f$''.}
%
\begin{figure}[ht!]
\centering \small
    \tikz{         
%         \node[const, above=of fr] (nfr) {$f_r$}; %
% 	\node[const, above=of nfr] (dfr) {\large $\mathbb{I}(d >0)$}; %

@@

\begin{equation*}\label{eq:m_v_f} \tag{\text{\en{product step}\es{paso productorial}}}

@@

\begin{equation*}\label{eq:m_f_v}  \tag{\text{\en{sum step}\es{paso sumatorial}}}

@@

\begin{equation*}\label{eq:marginal}  \tag{\text{marginal}}
p(v) = \prod_{h \in n(v)} m_{h \rightarrow v}

@@

\begin{equation*}\label{eq:gaussian_product} \tag{\text{\en{Gaussian product}\es{Producto de gaussianas}}}

@@

\begin{equation*}\label{eq:gaussian_division} \tag{\text{\en{Gaussian division}\es{División de gaussianas}}}

@@

\begin{equation*}\label{eq:integral_con_dirac} \tag{\text{\en{Dirac delta function}\es{Función delta de dirac}}}

@@

\begin{equation*}\label{eq:simetria} \tag{\text{\en{Gaussian symmentry}\es{Simetría de gaussianas}}}

@@

\begin{equation*}\label{eq:estandarizar} \tag{\text{\en{Gaussian standarization}\es{Estandarización de gaussianas}}}

@@

\begin{equation*}\label{eq:phi_norm} \tag{\text{\en{Derivative of the cumulative Gaussian}\es{Derivada de la gaussiana acumulada}}}

@@

\begin{equation*}\label{eq:phi_simetria} \tag{\text{\en{Symmetry of the cumulative Gaussian}\es{Simetría de la gaussiana acumulada}}}

@@

\en{The \texttt{Gaussian} class does most of the computation of the packages.}
\es{La clase \texttt{Gaussian} realiza la mayor parte del c\'omputo en todos los paquetes.}
%
\en{It is represented by two parameters, the mean (\texttt{mu}) and the standard deviation (\texttt{sigma}).}
\es{Se representa mediante dos par\'ametros, la media y el desv\'io estandar.}
% %
\begin{lstlisting}[backgroundcolor=\color{all},label=lst:N1_N2, caption={\en{Initialization of Gaussians distributions}\es{Inicialización de distirbuciones gaussianas}}, belowskip=-1.0 \baselineskip, aboveskip=-0 \baselineskip]
\end{lstlisting}
\begin{lstlisting}[backgroundcolor=\color{all}, belowskip=0.0 \baselineskip]
N1 = Gaussian(mu = 1.0, sigma = 1.0); N2 = Gaussian(1.0, 2.0)  
\end{lstlisting}
%
\en{The class overwrites the operators addition (\texttt{+}), subtraction (\texttt{-}), product (\texttt{*}) and division (\texttt{/}) with the main properties required to compute the marginal distributions in the TrueSkill Through Time model.}
\es{La clase sobreescribe los operadores suma (\texttt{+}), resta (\texttt{-}), producto (\texttt{*}) y divisi\'on (\texttt{/}) con las principales propiedades requeridas para computar las distribuciones marginales en el modelo TrueSkill Through Time.}
%
\begin{equation*} \tag{\texttt{N1 * N2}}
 \N(x|\mu_1,\sigma_1^2)\N(x|\mu_2,\sigma_2^2) \overset{\ref{multiplicacion_normales}}{\propto} \N(x|\mu_{*},\sigma_{*}^2)
\end{equation*}
%
\begin{equation*} \tag{\texttt{N1 / N2}}
 \N(x|\mu_1,\sigma_1^2)/\N(x|\mu_2,\sigma_2^2)  \overset{\ref{sec:division_normales}}{\propto} \N(x|\mu_{\div},\sigma_{\div}^2)
\end{equation*} 
%
\vspace{-0.3cm}
%
\begin{equation*} \tag{\texttt{N1 + N2}} \label{eq:suma_normales}
\begin{split}
\iint \delta(t=x + y) \N(x|\mu_1, \sigma_1^2)\N(y|\mu_2, \sigma_2^2) dxdy \overset{\text{\ref{suma_normales_induccion}}}{=} \N(t|\mu_1+\mu_2,\sigma_1^2 + \sigma_2^2)
\end{split}
\end{equation*}
%
\vspace{-0.5cm}
%
\begin{equation*} \tag{\texttt{N1 - N2}} \label{eq:resta_normales}
\begin{split}
\iint \delta(t = x - y) \N(x|\mu_1, \sigma_1^2)\N(y|\mu_2, \sigma_2^2) dxdy \overset{\text{\ref{suma_normales_induccion}}}{=} \N(t|\mu_1 - \mu_2,\sigma_1^2 + \sigma_2^2)
\end{split}
\end{equation*}

@@

\en{TrueSkill allows to estimate the skill of individuals even when they play in teams.}
\es{TrueSkill permite estimar la habilidad de los individuos incluso cuando estos juegan en equipo.}
%
\en{The model assumes that the team with the highest performance wins, $r = (t_i > t_j)$, where $t_i$ and $t_j$ represent the team performances.}
\es{El modelo supone que gana el equipo con mayor rendimiento, $r = (t_i > t_j)$, donde $t_i$ y $t_j$ representan los desempeños de los equipos.}
%
\en{In figure~\ref{fig:modelo_trueskill_2vs2} we show the factor graph.}
\es{En la figura~\ref{fig:modelo_trueskill_2vs2} mostramos la factorizaci\'on gr\'afica.} 
%
\begin{figure}[ht!]
  \centering
  \scalebox{.9}{
  \tikz{

@@

\en{Following the \ref*{eq:m_f_v} of the sum-product algorithm and the factorization of the model displayed at Figure~\ref{fig:modelo_trueskill_2vs2}, the messages from the skill factors $f_{s_i}$ to their variable $s_i$ are just the priors.}
\es{Siguiendo el~\ref*{eq:m_f_v} del \emph{sum-product algorithm} y la factorizaci\'on del modelo presentado en la figura~\ref{fig:modelo_trueskill_2vs2}, los mensajes de los factores de habilidad $f_{s_i}$ a su variable $s_i$ no son otra cosa m\'as que el priors.}
\begin{equation*}\label{eq:m_fs_s} \tag{\texttt{prior}}

@@

\en{The next message, the one that the variable $s_i$ sends to the performance factor $f_{p_i}$ is also the prior, in this case due to the~\ref*{eq:m_v_f} of the sum-product algorithm and the factorization of the model.}
\es{El siguiente mensaje, el que la variable $s_i$ le env\'ia al factor rendimiento $f_{p_i}$ tambi\'en es el prior, en este caso debido al~\ref*{eq:m_v_f} del \emph{sum-product algorithm} y la factorizacion del modelo.}

@@

\begin{equation*}\label{eq:m_fp_p} \tag{\texttt{performance()}}

@@

\begin{equation*} \label{eq:m_ft_t} \tag{\texttt{ta = p1 + p2}}

@@

\begin{equation*}\label{eq:m_fd_d} \tag{\texttt{d = ta - tb}}

@@

\begin{equation*}\label{eq:m_fr_r} \tag{\texttt{evidence}}

@@

\begin{figure}[ht!]

@@

\subsection{\en{The history of the Association of Tennis Professionals (ATP)}\es{Historia de la Asociación de Tenistas Profesionales (ATP)}} \label{sec:appendix_atp_code}

@@

%
\begin{lstlisting}[backgroundcolor=\color{python},caption={\en{\proglang{Python} code}\es{Código \proglang{Python}}},aboveskip=0.0 \baselineskip, belowskip=0.1cm]
Python
\end{lstlisting}
%
\begin{lstlisting}[backgroundcolor=\color{r},caption={\en{\proglang{R} code}\es{Código \proglang{R}}},aboveskip=0.0 \baselineskip, belowskip=0.1cm]
R
\end{lstlisting}

@@

\end{document}
